%!TEX root = ../dissertation.tex
% the acknowledgments section

%\newthought{Lorem ipsum dolor sit amet}, 

First I would like to thank my adviser, Sam Gershman, for taking me on, and teaching me the ropes, from the very ground up, of research in cognitive science. His open-mindedness, breadth of knowledge, and unending ideas, allowed me to cross discipline boundaries and develop a program of research that I am truly passionate about. His support and guidance have made this PhD a wonderful experience.

I would also like to thank several other people who have played advisory roles during my graduate studies. Noah Goodman for guiding my first foray into artificial intelligence and teaching me to think about how it interfaces with the study of human cognition. Josh Tenenbaum for his endless passion for and dedication to science, which has been an indispensable source of inspiration. Matt Botvinick for guiding me through a whole new way to think about cognition, and for introducing me to a whole new way to do science. Tom Griffiths whose lab I am very excited to be headed to next. As well as others (Fiery Cushman, Peter Dayan, Josh Greene, Todd Gureckis, Weiji Ma, Tim O'Donnell, Haim Sompolinsky, and Chris Summerfield to name a few) from whose ideas and advice I have drawn immense encouragement and inspiration.

I would like to thank my committee, Lakshinarayanan Mahadevan and Aravinthan Samuel for their help and feedback throughout this process. I am also grateful to the physics department at Harvard more broadly, in particular Jacob Barandes, Lisa Cacciabaudo, and Carol Davis, for their support.

Much of my work would have been impossible without my collaborators. I would like to thank Eric Schulz for being a constant (large positive constant) source of advice, ideas, encouragement, and silly jokes. Jane Wang and Zeb Kurth-Nelson for braving through the many trials and tribulations of cross-continental collaboration on cross-disciplinary topics. Demi Guo for being the best undergraduate research assistant anyone could hope for. Kevin Smith for his generosity, his attention to detail, and meticulous approach to science. I would also like to thank Peter Battaglia, Jeremy Bernstein, Silvia Chiappa, Edward Hughes, Rohan Hundia,  Jovana Mitrovic, Pedro Ortega, David Raposo, David Rolnick, Harrison Satcher, and Andreas Stuhlmüller.

Another wonderful aspect of my PhD has been the opportunity to interact with so many not only brilliant scientists, but also truly warm and inspiring people, through my work. I would first like to thank the members of and visitors to CCNLab at Harvard for being a great group of people to do science with, and to be around everyday: Rahul Bhui, Johannes Bill, Marcel Binz, Hayley Dorfman, Nick Franklin, Rob Geary, Yuval Hart, Hanna Hillman, Lucy Lai, John Mikhael, Edward Patzelt, Thomas Pouncy, Honi Sanders, Felix Sosa,  Momchil Tomov, Charley Wu and Jianqiao Zhu. I would also like to thank the broader computational cognitive science community for many stimulating discussions, and fun conference nights, in particular Kelsey Allen, Tyler Brooke-Wilson, Fred Callaway, Judy Fan,  Zenna Tavares, and Pedro Tsividis among many others. I have been very lucky to have had the opportunity to visit other institutions to do research during my PhD. I would like to thank the CoCoLab at Stanford University (in particular, Judy Fan, Robert Hawkins, and MH Tessler), as well as the Neuroscience team at DeepMind (in particular, Jess Hamrick, Kevin Miller, Kim Stachenfeld, Nick Watters, and Tina Zhu) for being very welcoming and for making these experiences productive and enjoyable.

I would also like to thank everyone whose friendship and support through the ups and downs of these early years in academia have made my PhD a memorable and gratifying experience. I would like to thank the physics crew: Ana, Anjalika, Anne, Baur, Bruno, Leo, Lisa and Victor, among many others, for the innumerable walks back. Bodhi, Priyank, Rushina, and Shibani for being my bedrock: I would not have survived the pains of growing up, and the confusions of a new country without you. Shamoyita, Bulbul-pishi, Susan \& John, and Pankaj-aunty for giving me respites into a `real home' so many times over, when I felt very far from my own.

I would also like to extend my gratitude to the many others I am sure to be forgetting who have contributed to my work and well-being during graduate school.

%I would also like to thank the many MTurk workers whose data this thesis is largely based on, as we

Finally I would like to thank my parents for their unconditional love and unwavering belief in me: none of this would be possible without them by my side. And Ian, for his kindness, his steadfast support, and his endless goofiness.

%Sam, Noah, Josh, Matt
%Tom
%Maha, Aravi
%
%Eric, Jane, Zeb, Silvia, Kevin, Demi, Jovana, David, Jeremy, Haim, Andreas.
%
%Nick, Rahul, Thomas, Yuval, Scott, John, etc
%
%Pedro Ortega, Nick Watters, Edward Hughes, Kevin Miller, Chris Summerfield, etc
%
%Robert Hawkins, Judy Fan, etc
%
%Bodhi, Priyank, Rushina, Shibani
%
%Bulbul, etc
%
%Anjalika, Ana, Anne, Bruno, Victor, Baur, Leo, Lisa, Zenna, Pedro, Tyler. 
%
%Parents, Ian



